How to \emph{not} write Virtual Machines for Dynamic Languages

### 2007-07-15 talk/dyla2007/dyla.tex
Final re-read of the paper, part 1.  More small fixes.

\item
\emph{Better GCs:} While this is obvious in theory, OO VMs tend to have a
higher memory overhead to start with.  For example, an instance of Sun's
Java VM which just loaded Jython consumes 34-42 MB of memory, while a
CPython process fits in 3-4 MB.

@@

An example where this mapping does not work very well is Prolog. While there
exist several implementations of Prolog on top of the JVM \cite{prologcafe}
\cite{InterProlog} and one on .NET \cite{psharp},
they are not particular efficient, especially when compared to good Prolog VMs
written in C. This is mostly because the Prolog execution model, which
involves backtracking and deep recursion, does not fit the JVM and .NET very
well. Therefore the Prolog implementations on top of OO VMs resort to models
that are quite unnatural both for the OO VM and for Prolog.

@@

The described proliferation of language implementations is a large problem for
language communities. Although most individual implementations exist for good
reasons, the sum of all of them and the need to keep them synchronized with the
reference implementations leads to a significant amount of duplicated work and division of
effort; this is especially true for open source languages which tend to evolve
quickly. At any one point in time some of the implementations will lag behind
which makes writing code which can work on all of the implementations harder.

Implementing a language on top of a OO VM has many advantages, so some
people propose the solution of standardizing on one particular OO VM to not have
to maintain implementations for several of them. While this would, in theory,
alleviate the problem it is unlikely to happen. On the one hand, many political
issues are involved in such a decision. On the other hand, deciding on a single
object and execution model would not be an equally good fit for all languages.

In the next section, we explore a different approach for implementing
dynamic languages that we hope is able to solve many of the problems of
implementing a language, in particular the problem of an explosion of the number
of implementations.

@@

The present paper proposes to approach the implementation of dynamic
languages from a meta-level: virtual machines for such languages should
not be written by hand, but generated automatically ``around'' an
interpreter playing the role of a high-level description of the language.  We
argue that this approach gives many of the benefits usually expected by
an implementer when he decides to target an existing object-oriented
virtual machine.  It also gives other benefits that we will describe --
mostly in term of flexibility.  Most importantly, it lets a
community write a single source implementation of the language, avoiding
the time-consuming task of keeping several of them synchronized.  The single
source can be used to generate either custom VMs for C-like
environments or interpreters running on top of OO VMs.  This makes it
practical to experiment with large changes to the language and with
entirely new languages, such as domain-specific languages, while at any
time being able to run the implemented language in a variety of
environments, from C/POSIX to the JVM to .NET.

@@

In meta-programming terms, the PyPy architecture is as follows:

@@

\item
We use a very expressive \emph{object language} (RPython -- an analyzable
subset of Python) as the language in which the complete Python
interpreter is written, together with the implementation of its
built-in types.  The language is still close to Python, e.g.  it is
object-oriented, provides rich built-in types and has automatic memory
management.  In other words, the source code of our complete Python
interpreter is mostly free of low-level details -- no explicit memory
management, no pieces of C or C-level code.

@@

\item
This meta-programming component of PyPy is called the \emph{translation
framework}, as it translates RPython source code (i.e. the full Python
interpreter) into lower-level code.  Its purpose is to add aspects to
and specialize the interpreter to fit a selectable virtual or hardware
runtime environment.  This either turns the interpreter into a
standalone virtual machine or integrates it into an existing OO VM.
The necessary support code -- e.g. the garbage collector when
targeting C -- is itself written in RPython in much the same spirit
that the Jikes RVM's GCs are written in Java \cite{JikesGC}; if needed, it is
translated together with the interpreter to form the final custom VM.
\end{itemize}

A detailed description of this translation process is beyond the scope of the
present paper; it can be found in \cite{pypyvmconstruction}.  The actual Python
interpreter of PyPy and the results we achieved by translating it to C, LLVM
\cite{LLVM} and .NET are described in \cite{architecture} \cite{translationdoc}.
These results show that the
approach is practical and gives results whose performance is within the same
order of magnitude (within a factor of 2 and improving) of the hand-written,
well-tuned CPython, the C reference implementation.  These figures do not
include the spectacular speed-ups obtained in some cases by the JIT compiler
described in section \ref{subsect:dynamic_compilers}.

>>>>

\item
\emph{Better GCs:} While this is obvious in theory, OO VMs tend to have a
higher memory overhead to start with.  For example, an instance of Sun's
JVM which just loaded Jython consumes between 34 and 42 MB of memory
(Linux/IA32), while on the same machine a CPython process fits in 3 to 4
MB.

@@

An example where this mapping does not work very well is Prolog. While there
exist several implementations of Prolog on top of the JVM \cite{prologcafe}
\cite{InterProlog} and one on .NET \cite{psharp},
they are not particularly efficient, especially when compared to good Prolog VMs
written in C. This is mostly because the Prolog execution model, which
involves backtracking and deep recursion, does not fit the JVM and .NET very
well. Therefore the Prolog implementations on top of OO VMs resort to models
that are quite unnatural both for the OO VM and for Prolog.

@@

The described proliferation of language implementations is a large problem for
language communities. Although most individual implementations exist for good
reasons, the sum of all of them and the need to keep them synchronized with the
reference implementation leads to a significant amount of duplicated work and division of
effort; this is especially true for open source languages, which tend to evolve
quickly. At any one point in time some of the implementations will lag behind;
for developers, this makes writing code which can work on all of the
implementations harder.

Implementing a language on top of a OO VM has many advantages, so some
people propose the solution of standardizing on one particular OO VM in order
to not have
to maintain implementations for several of them. While this would in theory
alleviate the problem it is unlikely to happen. On the one hand, many political
issues are involved in such a decision. On the other hand, deciding on a single
object and execution model would not be an equally good fit for all languages.

In the next section, we explore a different approach for implementing
dynamic languages that we believe is able to solve many of the problems that
implementers face, including the issue of the explosion of the number
of implementations.

@@

The present paper proposes to approach the implementation of dynamic
languages from a meta-level: virtual machines for such languages should
not be written by hand, but generated automatically ``around'' an
interpreter\footnote
{
We do not explore the possibility of starting from even
higher-level descriptions like declarative semantics; while interesting in
theory, there are many open issues that make it impractical so far, from
the question of whether it is feasible at all for large dynamic languages
to the issue of how to compile such a description to efficient code.
}
playing the role of a high-level description of the language.  We
argue that this approach gives many of the benefits usually expected by
an implementer when he decides to target an existing object-oriented
virtual machine.  It also gives other benefits that we will describe --
mostly in term of flexibility.  Most importantly, it lets a
community write a single source implementation of the language, avoiding
the time-consuming task of keeping several of them synchronized.  The single
source can be used to generate either custom VMs for C-like
environments or interpreters running on top of OO VMs.  This makes it
practical to experiment with large changes to the language and with
entirely new languages, such as domain-specific languages, while at any
time being able to run the implemented language in a variety of
environments, from C/POSIX to the JVM to .NET.

@@

In the terminology of meta-programming, the PyPy architecture is as follows:

@@

\item
We use a very expressive \emph{object language} (RPython -- an analyzable
subset of Python) as the language in which the complete Python
interpreter is written, together with the implementation of its
built-in types.  The language is still close to Python, e.g.  it is
object-oriented, provides rich built-in types and has automatic memory
management.  In other words, the source code of our complete Python
interpreter is mostly free of low-level details -- no explicit memory
management, no pieces of C or C-level code.  Any RPython code is also
valid Python code, so for testing and bootstrapping it can be run
directly on top of another Python implementation like CPython.

@@

\item
This meta-programming component of PyPy is called the \emph{translation
framework}, as it translates RPython source code (i.e. the full Python
interpreter) into lower-level code.  Its purpose is to add aspects to
and specialize the interpreter to fit a selectable virtual or hardware
runtime environment.  This either turns the interpreter into a
standalone virtual machine or integrates it into an existing OO VM.
The necessary support code -- e.g. the garbage collector when
targeting C -- is itself written in RPython in much the same spirit
that the Jikes RVM's GCs are written in Java \cite{JikesGC}; this
support code is
translated together with the interpreter to form the final custom VM.
\end{itemize}

A detailed description of this translation process is beyond the scope of the
present paper; it can be found in \cite{pypyvmconstruction}.  The actual Python
interpreter of PyPy and the results we achieved by translating it to C, LLVM
\cite{LLVM} and .NET are described in \cite{architecture} \cite{translationdoc}.
These results show that the
approach is practical and gives results whose performance is within the same
order of magnitude (within a factor of 2 and improving) of the hand-written,
well-tuned CPython, the C reference implementation.  These figures do not
include the spectacular speed-ups that PyPy obtains in some cases using the
JIT compiler generator
described in section \ref{subsect:dynamic_compilers}.
